\subsection{Volume II. The Congregation}

The second volume deals with the congregation. It contains 6 sections, or 25 pieces, 9 of which are unsigned editorial articles from Sverdrup’s pen. For local reasons, almost all these treatises and contributions were composed in a popular form; the presentation is always clear and transparent; but it lies in the nature of the matter that not a little requires labor on the reader’s part. Here too the rule applies: “To the one who has, more shall be given.” The one who comes to the book with questions will receive answers.

It is not our intention to enter into the individual articles. We shall mention most of them and attach remarks to such as seem to us to have a particular significance for the understanding of the ecclesiastical development among us.


\subsubsection{First Section}

1. “What Is at Stake?”, written as early as 1877, with reference to the controversy carried on between the Synod and the Conference, shows that even a whole generationbefore Sverdrup ended his labors, it was the congregation that was at issue: “Our definite conviction is this, that in this controversy it is the congregation that is at stake.”

2. “The Holiness of the Congregation” states clearly that the holiness of the congregation, and the question bound up with it, “What is a congregation?”, constitutes “the real main question in the Free Church.” It should be noted, first, that the article was written in 1889, the year before the union, and many years before the Lutheran Free Church was formed; second, that Sverdrup was opposed to the “unsound views” for which Pastor Trandberg and his Reformed compatriots were spokesmen.

It is emphasized that there may be a congregation without organization or formal constitution; for the congregation in Jerusalem was not organized; further, that the empirical congregation is still truly a congregation, even though many of its members are hypocrites, dead, sleeping, unbelieving, and ungodly. Sverdrup’s position here—as it always was—is in agreement with Article 7 of the Augsburg Confession, which depicts the congregation according to its idea, true essence, its invisible-visible form—and with Article 8, which depicts the congregation according to its historical-empirical appearance, a mixture of hypocrites and evil men with the believers, who form the congregation’s real core.

3. “The Liberation of the Congregation,” 1896, points out that it was the new direction in the Conference that “held back,” so that “the congregation’s being kept in a state of dependency” did not proceed so rapidly, and that the old direction found staunch allies in the union, so that the United Church in the course of a few years left “all the other bodies far behind in ecclesiastical corporatism and tyranny over the congregations.” The author lays the blame for this partly upon the State Church upbringing. “Many times that which in the State Church was done by compulsion is here among us imitated by habit and custom. Custom among the people is a stronger power than the law for compelling men, and the old habits reign long after legal compulsion has ceased, especially when many zealous pastors labor with all their might to maintain the old habit and custom, that all shall belong to the congregation” (p. 19). Among other things, S. emphasizes the necessity of lay activity and points out that in the Norwegian State Church there is quite great freedom for lay activity because the law protects it, and that there was far greater freedom for lay activity there than in most “majority congregations.”

Distinctions of rank and class, and clerical pomposity, must go.

4. “The Quickening of the Congregation,” 1898, includes, among other things, substantial material on Luther’s understanding of church order.

\subsubsection{Second Section}

1. “Which Church Polity Is Most Fitting for the Evangelical Lutheran Church?”, 1869. This is perhaps the oldest literary contribution from Sverdrup’s hand that we now possess. It is interesting because it gives us an insight into what a theological student, 20 years old, could say about church polity. One will be astonished that Sverdrup at so young an age could grasp a question of church polity with such deep interest as this treatise bears witness to, and still more at the ability with which it is handled. As far as I can see, S. is here the spokesman for Rudelbach’s ideas, in that he pleads for consistories and synods, but disapproves of congregations of the independent type.

In our country Sverdrup was, from first to last, the defender of the independent congregations. In a free country, he held that the episcopal, the presbyterial, the consistorial polity that one had in the Lutheran lands of Europe stood far behind the congregational form of polity. The Free Church was therefore built upon a congregationalist basis. That was something new—at any rate in the history of the Lutheran Church. In Europe too eyes have been opened to the superiority of the polity of a free congregation. None less than the clear-sighted and learned church historian Loofs at the University of Halle says: “Who knows whether the congregationalist church polity does not also have its future among us, when one day the State Churches in the old world collapse.”

2. “The Free Church Community,” 1877–78. This treatise, of 46 pages, I regard as the most important in the whole book. It falls into six subdivisions: The Congregation, The Local Congregation, The Office in the Free Congregation, The Pastoral Office and Lay Activity, The Diaconal Office in the Congregation, The Fellowship of the Congregations. Here we find answers to questions such as: how are we to found congregations? what does it mean “to compel them to come in”?

It is the Word, says S.—not outward compulsion or inducements or party spirit—that shall gather a congregation. “Even among the Norwegian emigrants, who indeed had all once been received into God’s Church through Baptism, it would have been far better if one had never organized any congregation or gathered congregational members in any other way or by any other means than by the preaching of God’s Word, publicly and privately. For even baptized persons are many a time, alas, in such a condition that it is better both for themselves and for the congregation that they ‘stand outside’ and are the object of the congregation’s influence in that way, than that they should be inside and perhaps there sleep the deepest sleep, precisely because they are ‘in the congregation.’”

After dwelling somewhat upon the all too hasty admissions into the congregation, and the dangers bound up with church discipline and exclusion, Sverdrup lays the weight upon this: that the congregation must be gathered by God’s Word, and that the constr