% FIXME: Refine these ideas: Fail forward. Faith alone, but faith is strengthened by work and a call to action. Give the doubting member work to do. Give them work because God has already placed them within obligations of daily bread, neighbor, family, congregation, and creation. In that work, faith is often exercised before it is fully understood.


\subsection{How Can We Hear Sverdrup, Evjen, and The Editors?}

%Let's start with a quote from Prof. Elias Harbo (first president of the LFC). 

%\begin{quote}

%The holy cause of the freedom of the congregation had, accordingly, still no spokesperson among us. Yet the day was dawning, the day when God's light in the world concerning the congregation would be set on a stand and shine out over the Norwegian Lutheran Church people in this country.

%\end{quote}

%This type of language would make Evjen very uncomfortable as it places professors Georg Sverdrup and Sven Oftedal high on a pedestal. From that lofty position, there are few who are willing to ask questions; after all, the message is given. Yet, that is precisely what Evjen was all about. We are called to wrestle with Evjen as he has something to teach us about history, faith, and ugly ways polity impacts us today.     
\textit{Apologetics begins with a hymn.} 

We cannot understand Sverdrup merely by reading Sverdrup, because we no longer have the common frame of reference. As evidence, consider this clause from P1: ``Kingdom of God on earth.'' I ask you to take it seriously with all the implications applicable to the congregation and individual members operating within and beyond the church.

This is a good time to stop and find an old hymnal, as our hymns are downstream of theology. They are not neutral; they carry the emotion, prayer, and outlook. They also reflect a sense of mission. In modern language, we would say that hymns encode a call to action. This allows us to gauge the theological temperature of a period without drowning in historical detail.     

To fully understand Sverdrup's time period, immerse yourself in the Kingdom described by these hymns:

\begin{itemize}
    \item \emph{Battle Hymn of the Republic}
	\item \emph{Stand Up, Stand Up for Jesus}
	\item \emph{We’ve a Story to Tell to the Nations}
	\item \emph{Onward, Christian Soldiers}
	\item \emph{Bringing in the Sheaves}
	\item \emph{From Greenland’s Icy Mountains}
    \item \emph{Work, for the Night Is Coming}
\end{itemize}

There is an incredible optimism woven into these hymns, an American heat whose supreme confidence borders on sinful pride. They all sing in the same triumphant key about the Kingdom coming to all the earth. Do we dare take the next step and declare that the "Kingdom of God on earth" is here embodied in the congregation and its faithful members? Is this not the force pressing P1 forward? 


\begin{quote}
According to the Word of God, the congregation is the right form of the Kingdom of God on earth.
\end{quote}

You can read more about this uniquely American church history era under the category of postmillennialism. You will find the power best exemplified in the refrain of ``We’ve a Story to Tell to the Nations:''

\begin{quote}
For the darkness shall turn to dawning,\\
and the dawning to noonday bright,\\
and Christ's great kingdom shall come on earth—\\
the kingdom of love and light.
\end{quote}

This is the providential calling to the congregation as well as individual duty, urgency, and frankly, the unquestioned march driving the church's mission forward. It's also mixed up with American manifest destiny, and even technology as factories, agriculture, and steam power reshape life. 

% Removed Harbo

Understand that this is the empowering water that surrounded Sverdrup, Oftedal, Evjen, along with the Congregationalists in the mid-to-late 1800s up until about World War 1. Like a fish in water, I'm not convinced they could see it. They certainly would not have used the term \emph{postmillennialism}, instead preferring to describe the life of the Norwegian immigrant in America. Yet, it is this same American theological current that carried ``John Brown's Body'' to your hymnal with the words: 

\begin{quote}
Glory! Glory! Hallelujah!\\
Glory! Glory! Hallelujah!\\
Glory! Glory! Hallelujah!\\
His \st{soul} truth is marching on.
\end{quote}


Do not write off Sverdrup and Oftedal as postmillennial thinkers. To the contrary, they offer something profound that calls to us today. The problem is that we are a century removed. The language has changed and the shared metaphors with their deadly serious polemical force have softened so much that we only hear the echo. 

We can clearly see the postmillennial thinking because we are at least two hymnal generations removed. I'd argue that the first shift occurred around the 1950s, with a de-emphasis on anything that looked like the Protestant vocational work ethic, striving, or moral progress carried beyond Sunday morning. Hymns such as \textit{Work, for the Night Is Coming} became uncomfortable for many Lutherans. The latest hymnal generation emerged around the turn of this century with an ever-growing emphasis on comfort and assurance. The unyielding hammer of judgment---that we are all sinners---has become a hidden footnote. The evidence is written in the hymnals, or in some cases, the abandonment of hymnals altogether. 

The opening hymns set the stage. They remind us of the American confidence into which P1 was born. Yet, it would be a mistake to nostalgically return to those times. Instead, we need something stronger, more disciplined, and biblically grounded. Let's continue using a ship analogy with a modern understanding of vocation. But before we do, let's reset our ear to the confessional register with one of Luther's greatest hymns. Ponder the embodied wisdom of this hymn. The wisdom is inseparable from the anchor of the Kingdom of God on earth. We cry from the depths. Our lives in the way of the old Adam are unmanageable. We cannot stand upright before the Judge who remembers every sin. Yet, there is our Shepherd to lead the way.     

\pagebreak

\textit{Out of the Depths I Cry to Thee}:

\begin{quote}

1 From depths of woe I cry to Thee,\\
Lord, hear me, I implore Thee.\\
Bend down Thy gracious ear to me;\\
my prayer let come before Thee.\\
If Thou rememb'rest every sin,\\
if naught but just reward we win,\\
could we abide Thy presence?\\

2 Thy love and grace alone avail\\
to blot out my transgression;\\
the best and holiest deeds must fail\\
to break sin's dread oppression.\\
Before Thee none can boasting stand,\\
but all must fear Thy strict demand\\
and live alone by mercy.\\

3 Therefore my hope is in the LORD\\
and not in mine own merit;\\
it rests upon His faithful word\\
to them of contrite spirit\\
that He is merciful and just;\\
this is my comfort and my trust.\\
His help I wait with patience.\\

4 And though it tarry till the night\\
and till the morning waken,\\
my heart shall never doubt His might\\
nor count itself forsaken.\\
Do thus, O ye of Adam's seed,\\
ye of the Spirit born indeed;\\
wait for your God's appearing.\\

5 Though great our time and sore our woes,\\
His grace much more aboundeth;\\
His helping love no limit knows,\\
our utmost need it soundeth.\\
Our Shepherd good and true is He,\\
who will at last His people free\\
from all their sin and sorrow.\\
\end{quote}

\pagebreak


\subsection{A Model of the Congregation}

\textit{P1 is exposed by our disobedience.}

The model described in this section is required to visualize the dangers described by Evjen and by The Editors. Like many Lutheran ideas, it is held in tension---between the ditches as Luther might say. Evjen asks us to seriously consider the visible form and order that hold the congregation together; The Editors ask us to consider the living Spirit within those structures. Eliminate either side and the congregation collapses. 
 
To better understand this world, I offer a model of the visible congregation as a ship on a mission at sea. This model allows us to better understand the water in which the visible congregation is sailing. Evjen was writing about Sverdrup in the decade following his untimely death. We are reading Evjen and Sverdrup a century later. I suspect Evjen was too close to the fire and we are so far away that we cannot distinguish the smoke through the fog. The model provides a conceptual framework that makes it easier to visualize the congregation. Conversations are easier, and operational failures become easier to describe and harder to deny.

Keep in mind that the ship is my metaphor and, to the best of my knowledge, not used by Sverdrup, Oftedal, or Evjen. The ship is a natural metaphor for me as I served on icebreakers and a three-masted barque. I've sailed far north of Alaska and south to Antarctica. Let me share the first lesson of sailing. A ship is built for a crew, not an individual. Crewmembers are specialized and handle the ship’s daily work according to their gifts and training. The ship and everyone on board depend on the faithful work of the individual in service to the ship in service of the mission. 
     
Recall that P1 does not fit cleanly within dogmatics, ecclesiology, or polity boundaries. Yet our understanding requires that we anchor each statement within that framework. For the ship, dogmatics declares what we believe from Scripture, ecclesiology describes the nature of the ship and crew, and polity describes the crew's operational obligations. I've added what I believe is the correct category for each statement. I encourage you to test my classifications, especially when more than one term applies.  

Quoted material in this list is contained within the Guiding Principles (See front material). 


\paragraph{The Ship is Given by Christ}
\begin{itemize}

\item Dogmatics: The ship is a gift from Christ originating at Pentecost in the upper room.

\item Ecclesiology: The ship's hull marks the visible and bounded form of the congregation. It defines the boundary between internal Christian obligation and mission to the outside world.

\item Ecclesiology: There are many ships of many designs. Yet none are warships wielding the sword. All have a mission from Christ. 

\item Ecclesiology: Within the ship is a living and visible congregation that acts. 

\end{itemize}



\paragraph{The Ship’s Power and True Center}
\begin{itemize}

\item Dogmatics and Ecclesiology: The Word is preached clearly so that all hands can point the way from old Adam to the promise of the Cross. 

\item Dogmatics: The ship is powered by the Word and sacraments. All crewmembers are marked with the cross at baptism. The crew kneels for the body and blood of Christ given and shed for you. Yet the sacraments are not hidden in the engine room. They are visible and sacred acts received and witnessed by all crewmembers.

\item Dogmatics: Justification was purchased at the cross. Sanctification is the Spirit's work. In the Reformation tradition, we confess that faith alone justifies, but faith is never alone. Consequently, we fear a passive crew as the ship is in peril.

\end{itemize}



\paragraph{The Crew as a Living Congregation}
\begin{itemize}

\item Polity: The crew learn about each other over coffee and cookies. This is not an optional exercise as it is one of the primary ways that members of the congregation become known to one another. The resulting prayers are real as the crew cares enough to hear how others are suffering. 

\item Ecclesiology: An individual’s contributions changes with time. We all have seasons. God willing, we pass from child, to youth with strength and vigor, to wisdom, to autumn, to fading strength. Gifts change, yet we we remain part of the body. All hands are required at every stage.
 
\item Ecclesiology: A healthy ship is in a constant state of training to reinforce both the spiritual as well as vocational skills of the crew.  
 
\item Polity: Bear ye one another's burdens, and so fulfil the law of Christ.

 \end{itemize}



\paragraph{Order, Watchstanding, and Lay Responsibility}
\begin{itemize}

\item Polity: Each ship must have order or it will sink, dash against the rocks, or burn to the waterline, leaving the crew stranded. When the storm strikes, the crew must be ready to respond reflexively.

\item Ecclesiology: Each member is ``constrained by the Spirit of God and by the privileges of Christian love'' to freely exercise their spiritual gifts. 

\item Polity: The active use of all hands is essential. However, there are select positions that must be filled to maintain a healthy congregation. There must be leadership along with serious lay members to strengthen leadership as steel on steel. Watchstanders are required to identify the obstacles and the opportunities: Sunday school teachers, Bible school leaders, council members, treasurers, and perhaps more importantly, caring members who extend a welcoming hand to visitors or to those who cannot attend.  

\item Ecclesiology and Polity: The crew must sound the ship and tap the pipes to retain the knowledge of what keeps the ship afloat and running. Faithful laity must understand the ship well enough to inspect and help repair as necessary. 

\item Polity: All hands are alert; every faithful member is a watchstander. The alarm must be sounded when the ship is in danger.    

\end{itemize}



\paragraph{Mission Beyond Sentiment}
\begin{itemize}

\item Ecclesiology: The ship, as a visible congregation, is on a mission with a calling from Christ. The congregation is composed of individual sinners gathered together as the ship's crew. The ship is the visible structure in which Christ's work is carried out. 

\item Ecclesiology and Polity: The ship gathers sinners who would otherwise drift without purpose, community, and meaning. They are welcomed long before their station is clear. Brotherhood, correction, responsibility, and training are given so that each may bear life's load under God's rule in both kingdoms. 

\item Ecclesiology and Polity: The ship carries memory as obligation. The boundaries and mission are inherited from those who can no longer stand the watch. They are honored through faithful action to maintain the ship and mission on Christ's given course. 

\end{itemize}




\paragraph{Danger, Failure, and the Anti-Sentimental Guardrail}
\begin{itemize}

\item Dogmatics: This is not a naturally healthy ship. We are all sinners full of pride, resentment, and hypocrisy. 

\item Dogmatics: Satan is whispering to the crew, and the real world presses in with political, societal, and economic constraints. There are also openly hostile enemies to Christianity. 

\item Ecclesiology: This is not a safe ship. It's not some sentimental ark to weather the storm. It's an operational ship dependent on the shared responsibility of the crew to perform the ship's business while taking care of all members. There is an old saying that ``a ship in port is safe, but that's not what a ship is for.''

\item Ecclesiology: A congregation may appear safe when it avoids mission, repair, conflict, responsibility, and pays the pastor until it closes its doors. But that's not what a congregation is for. Or as Sverdrup once stated, a congregation may appear to have all its members, but it may be missing the most important thing: life. 

\item Dogmatics and Ecclesiology: Being a member of the crew does not, in itself, imply that we are saved. This is not works-righteousness. You will not be saved by working harder, longer, or by advancing in station aboard the ship. Not even the captain is automatically saved. However, in the same breath, you are called to work ``as far as its abilities and power permit.'' 

\end{itemize}



In summary, this ship is used to visualize the Pauline language of Ephesians 4:16:  

\begin{quote}

From him the whole body, joined and held together by every supporting ligament, grows and builds itself up in love, as each part does its work.

\end{quote}
 

\subsection{A Broken Model of the Congregation}


\textit{The danger is not that P1 is too difficult to understand, but that it is too costly to believe.}

The ship metaphor works. We can clearly see that the congregation acts. It carries the gifts of Christ to the world. The congregation and every member within depend on the faithful work of each individual in service to the congregation in service of the mission in Christ.

The individual responsibility to the congregation is clear. Yet, the model breaks precisely at the level of individual responsibility. Let me explain with another hard-learned sailor's lesson, ``You represent the ship. Act accordingly as you go ashore.'' Stated another way, if you are a member of the visible congregation that confesses itself to be the ``Kingdom of God on earth,'' then you have obligations. It's an uncomfortable position for the Lutheran and for the American Christian in general, however, the implications are unavoidable. 

Even more uncomfortable is the fact that the relationship runs both ways. The congregation has an obligation to those who go ashore. It does not end with confirmation and magically restart when marriage and children make the individual visible again. If we profess that the congregation is the Kingdom of God on earth, people are not left to drift. The congregation's vocation of brotherhood, correction, responsibility, and training extends beyond the hull. To be blunt, the nice polite sailor abandons the crew by ignoring fire, flooding, or a drowning shipmate. It is an abdication of duty.      

Let's refine our definitions to include Luther's concept of vocation with assistance from Veith:
  
 \begin{itemize}
    \item \textbf{Dogmatics:} What we confess because Scripture teaches it.

    \item \textbf{Ecclesiology:} What we believe Scripture teaches about the nature of the congregation.

    \item \textbf{Polity:} How we organize and operate the congregation.

    \item \textbf{Vocation:} How an individual operates outside the ship and in the sinful world in terms of work, family, neighbor, and civic obligations. The sovereign individual does not escape obligation while ashore. Freedom is rightly ordered through the service of daily bread. At the same time, individuals within the congregation have a vocation toward those who go ashore.  

\end{itemize}

Just as the ship has a mission, so too does the individual. We are anchored in the Principles, ``every individual believer is constrained by the Spirit of God and by the privileges of Christian love...'' We could use Pauline language and talk about a leg that wants to leave the body, as if the severed leg could live or the body could continue its mission unharmed. Christ gives the authority and Paul gives us the body. The ship helps us see the body at work.    

Principle One is not merely an academic problem as framed by Evjen, but an existential vocational problem. At its core it demands something from every member of the crew while on board and ashore. Remember that the congregation is a gift from Christ. It's an operational support vessel with a single course toward the Cross. We are called to carry Christ in all that we do in the congregation, family, neighbor, and yes, even nation. The Spirit and life are given. However, these gifts must be developed by all hands for the glory of God and the necessity of daily bread. Sverdrup would say:

\begin{quote}
By congregation we mean a gathering of active Christians, who use the means of grace and the gifts of grace for their own and their fellow men’s salvation and blessedness. Where such a gathering of active Christians is found, the question does not become one of binding and controlling lay activity and other gifts of grace, but much more of giving them room and freedom, occasion and encouragement, so that they may come to the greatest flourishing and the most effective use.\footnote{See Sverdrup’s full article, “The Congregation and Lay Activity}
\end{quote}

In modern language: do not sin against vocation.

If we are unwilling to develop and use these gifts ``as far as its abilities and power permit'' the congregation fails. ``Spirit and life'' is not a revival statement about \textit{Jesus Loves Me}, but a life-and-death question about the duty performed by the individual members of the congregation. In biblical terms, P1 is a declaration that works follow faith: we are saved by grace through faith, ``not a result of works,'' yet “created in Christ Jesus for good works” (Eph. 2:8–10). A serious acceptance of the ``Kingdom of God on earth'' demands no less.  

This may seem like an insurmountable problem, yet it has been with us all along as Principle 12:

\begin{quote}
Every free congregation, as well as every individual believer, is constrained by the Spirit of God and by the privileges of Christian love to do good and to work for the salvation of souls and the quickening of the spiritual life, as far as its abilities and power permit. Such free spiritual activity is limited neither by parish nor by synodical boundaries.
\end{quote}

And finally we arrive at Sverdrup's and Evjen's core message concerning a living and visible congregation. This is a calling. It acts. There is room for all of God's gifts and work for every hand. 

Can we bear the weight of our earthly vocation? Will we accept P1 not as a sentimental statement of spirit and life but as a call to act as serious members of the body of Christ? Evjen calls to us:

\begin{quote}
The question—the real one—becomes: Does one stand in living accord with the future of God’s kingdom, and is one willing to let oneself be used for the kingdom’s advance, to labor for the fulfillment of the prayer: hallowed be your name; your kingdom come; as in heaven so also on earth—your will be done!”\footnote{Installment 11: “Folkebladet” May 22, 1918}
\end{quote}  


\begin{quote}
“It is possible to bury the principles and get a peace that is as silent as the grave’s.”\footnote{Installment 9: “Folkebladet” May 08, 1918}
\end{quote} 


\begin{quote}
“My language has been the bracing language of duty.”\footnote{Installment 9: “Folkebladet” May 08, 1918}
\end{quote} 


%Can P1 exist as a living confession without the fruits of ``every single man and woman who is called in Christ's name?'' 