\subsection{Disclaimer}

This volume is an independent scholarly and technical project. It is not affiliated with, endorsed by, or issued under the authority of any church body, academic institution, or other organization, including but not limited to the Association of Free Lutheran Congregations (AFLC), the Sverdrup Society, Augsburg Seminary, or any related bodies.

The texts presented herein are historical documents. The English translations, transcriptions, and editorial materials are supplied solely for educational and research purposes. They do not claim doctrinal, ecclesiastical, institutional, or critical authority.

The English translations in this volume were produced with the aid of artificial intelligence-based language tools operating under detailed scholarly instructions. Editorial responsibility consists of methodological design of the AI translation prompts, critical oversight, source comparison, translation review, and presentation. An honest assessment reveals that shards of Aaron Dahlen are present in the translation; none of this was a simple click of the button.

The original source texts remain definitive.


\section*{Editorial Note}

This document is a working translation and reconstruction from a source that is in poor physical and typographic condition. Some passages remain uncertain owing to faded print, damaged pages, archaic spelling, irregular orthography, and the inherent limitations of OCR and other machine-assisted recovery methods; such tools can assist transcription, but they cannot by themselves guarantee correct readings in difficult witnesses.

Readers should therefore treat this as a provisional scholarly working text rather than a final critical edition. The project is maintained openly on GitHub and constructed in \LaTeX, so that the text, formatting, annotations, and editorial decisions may be revised transparently over time.

Corrections, source checks, and suggested improvements are welcome through the project repository on GitHub, where revisions and editorial decisions may be tracked transparently.


\url{https://github.com/apdahlen/Evjen}
